---
title: "Lotka-Volterra Competition Analysis"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis fits a two-species Lotka-Volterra competition model to paired
rotifer (*Habrotrocha rosa*) and protist (*Tetrahymena*) growth curves from
the competition experiment. For each competition replicate, four parameters
are estimated:

- `a_RR`: rotifer intraspecific crowding coefficient
- `a_RP`: per-capita effect of protists on rotifers (interspecific)
- `a_PP`: protist intraspecific crowding coefficient
- `a_PR`: per-capita effect of rotifers on protists (interspecific)

Intrinsic growth rates (`r_R`, `r_P`) are fixed at values estimated from
single-species monocultures, varying by treatment combination.

The ODE system fit is the Mallet (2012) form:

$$\frac{dN_R}{dt} = N_R (r_R - \alpha_{RR} N_R - \alpha_{RP} N_P)$$
$$\frac{dN_P}{dt} = N_P (r_P - \alpha_{PP} N_P - \alpha_{PR} N_R)$$

# Setup

```{r libraries}
library(deSolve)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(stringr)

# Reproducibility
set.seed(42)
```

# Load data

```{r load_data}
# EDIT: point this to wherever you have exp2_growthcurves saved
# load("path/to/exp2_growthcurves.RData")

# Assuming exp2_growthcurves is already loaded in your environment:
gc <- exp2_growthcurves

# Sanity check structure
str(gc)
table(gc$species, gc$competition)
table(gc$treatID)
```

# Data prep

Convert counts to per-mL densities and reshape so each row is a single
timepoint with both species' densities.

```{r data_prep}
# Confirm `count` is already converted to per-0.1-mL (counts column should
# equal fullSlide where fullSlide is non-NA, or singlePass * 6.8 where it is).
# This check is sanity only; if it fails, the conversion needs to happen here.

check_conversion <- gc %>%
  mutate(
    expected = case_when(
      !is.na(fullSlide)  ~ as.numeric(fullSlide),
      !is.na(singlePass) ~ singlePass * 6.8,
      TRUE               ~ NA_real_
    ),
    diff = abs(count - expected)
  ) %>%
  summarise(
    max_diff = max(diff, na.rm = TRUE),
    n_mismatches = sum(diff > 0.01, na.rm = TRUE)
  )
print(check_conversion)

# Counts in `count` are per 0.1 mL; convert to per mL
gc <- gc %>%
  mutate(density_per_ml = count * 10)

# Keep only competition treatments (treatID ending in "RP")
comp_data <- gc %>%
  filter(str_ends(treatID, "RP"))

# Reshape so each row has both species' densities at each timepoint
comp_wide <- comp_data %>%
  select(day, evolvedTemp, currentTemp, treatID, repNum, species, density_per_ml) %>%
  pivot_wider(names_from = species, values_from = density_per_ml,
              names_prefix = "N_") %>%
  rename(N_R = N_rotifer, N_P = N_protist) %>%
  arrange(treatID, repNum, day)

head(comp_wide)
```

# Set up r values

These come from monoculture fits. Hourly rates converted to per-day.

```{r r_values}
# Rotifer r by evolved x current temperature (hourly -> daily)
r_rotifer_lookup <- tribble(
  ~evolvedTemp, ~currentTemp, ~r_R_hourly,
  "25", "25", 0.0203,
  "25", "30", 0.0332,
  "30", "25", 0.0193,
  "30", "30", 0.0294
) %>%
  mutate(r_R = r_R_hourly * 24)

# Protist r by current temperature only (hourly -> daily)
r_protist_lookup <- tribble(
  ~currentTemp, ~r_P_hourly,
  "25", 0.0695,
  "30", 0.214
) %>%
  mutate(r_P = r_P_hourly * 24)

# Single-species alphas (monoculture). Also hourly; will convert as needed.
# Used as initial guesses for fitting, not as fixed values.
a_rotifer_lookup <- tribble(
  ~evolvedTemp, ~currentTemp, ~a_RR_hourly,
  "25", "25", 0.000305,
  "25", "30", 0.000436,
  "30", "25", 0.000318,
  "30", "30", 0.000435
) %>%
  mutate(a_RR_init = a_RR_hourly * 24)

a_protist_lookup <- tribble(
  ~currentTemp, ~a_PP_hourly,
  "25", 0.0000675,
  "30", 0.000315
) %>%
  mutate(a_PP_init = a_PP_hourly * 24)

print(r_rotifer_lookup)
print(r_protist_lookup)
print(a_rotifer_lookup)
print(a_protist_lookup)
```

# LV model definition

```{r lv_model}
# Two-species Lotka-Volterra in Mallet form
lv_model <- function(t, state, parms) {
  N_R <- state["N_R"]
  N_P <- state["N_P"]

  dN_R <- N_R * (parms["r_R"] - parms["a_RR"] * N_R - parms["a_RP"] * N_P)
  dN_P <- N_P * (parms["r_P"] - parms["a_PP"] * N_P - parms["a_PR"] * N_R)

  list(c(dN_R = dN_R, dN_P = dN_P))
}

# Simulate the LV system given parameters and timepoints
simulate_lv <- function(params, times, N_R0 = 40, N_P0 = 80) {
  state0 <- c(N_R = N_R0, N_P = N_P0)
  out <- tryCatch(
    ode(y = state0, times = times, func = lv_model, parms = params,
        method = "lsoda"),
    error = function(e) NULL
  )
  if (is.null(out)) return(NULL)
  as.data.frame(out)
}

# Negative log-likelihood for one replicate (Gaussian errors on log scale)
# Estimating four parameters: a_RR, a_RP, a_PP, a_PR (r values fixed)
nll_lv <- function(log_params, data, r_R, r_P, N_R0 = 40, N_P0 = 80) {
  # Exponentiate to enforce positivity
  params <- c(
    r_R  = r_R,
    r_P  = r_P,
    a_RR = exp(log_params[1]),
    a_RP = exp(log_params[2]),
    a_PP = exp(log_params[3]),
    a_PR = exp(log_params[4])
  )

  times <- sort(unique(data$day))
  sim_times <- if (min(times) > 0) c(0, times) else times
  sim <- simulate_lv(params, times = sim_times,
                     N_R0 = N_R0, N_P0 = N_P0)
  if (is.null(sim)) return(1e10)

  # Rename simulation outputs to avoid column-name collision on join
  sim <- sim %>%
    rename(sim_N_R = N_R, sim_N_P = N_P)

  # Match simulation timepoints to data
  matched <- data %>%
    left_join(sim, by = c("day" = "time")) %>%
    filter(!is.na(sim_N_R), !is.na(sim_N_P))

  if (nrow(matched) == 0) return(1e10)

  # Penalize numerical failures (NaN, Inf) explicitly
  if (any(!is.finite(matched$sim_N_R)) || any(!is.finite(matched$sim_N_P))) {
    return(1e10)
  }

  # Log-space residuals; add 1 to avoid log(0). Rotifer and protist residuals
  # contribute equally (could be weighted by species-specific variance).
  resid_R <- log(matched$N_R + 1) - log(pmax(matched$sim_N_R, 0) + 1)
  resid_P <- log(matched$N_P + 1) - log(pmax(matched$sim_N_P, 0) + 1)

  ssq <- sum(resid_R^2, na.rm = TRUE) + sum(resid_P^2, na.rm = TRUE)
  if (!is.finite(ssq) || length(resid_R) == 0) return(1e10)
  ssq
}
```

The simulation outputs are renamed to `sim_N_R` and `sim_N_P` before joining
to avoid column-name collision with the observed `N_R` and `N_P`. If your
data uses different column names for observed densities, update the `rename`
call and the residual computation accordingly.

# Fit a single replicate (sanity check)

Before batch fitting, run on one replicate to verify the pipeline works.

```{r fit_one_replicate}
# Pick one replicate to fit as a sanity check
test_rep <- comp_wide %>%
  filter(treatID == "E25HRP", repNum == "1") %>%
  arrange(day)

print(test_rep)

# Get r values for this treatment
r_R_test <- r_rotifer_lookup %>%
  filter(evolvedTemp == as.character(test_rep$evolvedTemp[1]),
         currentTemp == as.character(test_rep$currentTemp[1])) %>%
  pull(r_R)

r_P_test <- r_protist_lookup %>%
  filter(currentTemp == as.character(test_rep$currentTemp[1])) %>%
  pull(r_P)

# Initial parameter guesses (on log scale)
# Use monoculture alphas as starting points for self-suppression terms
a_RR_init <- a_rotifer_lookup %>%
  filter(evolvedTemp == as.character(test_rep$evolvedTemp[1]),
         currentTemp == as.character(test_rep$currentTemp[1])) %>%
  pull(a_RR_init)

a_PP_init <- a_protist_lookup %>%
  filter(currentTemp == as.character(test_rep$currentTemp[1])) %>%
  pull(a_PP_init)

# Initial guesses for interspecific terms: start small
log_init <- log(c(a_RR_init, a_RR_init * 0.5, a_PP_init, a_PP_init * 0.5))

fit_test <- optim(
  par = log_init,
  fn = nll_lv,
  data = test_rep,
  r_R = r_R_test,
  r_P = r_P_test,
  N_R0 = 40,
  N_P0 = 80,
  method = "Nelder-Mead",
  control = list(maxit = 5000, reltol = 1e-8)
)

print(fit_test)
cat("\nEstimated parameters:\n")
est_params <- exp(fit_test$par)
names(est_params) <- c("a_RR", "a_RP", "a_PP", "a_PR")
print(est_params)
```

# Visualize the fit

```{r visualize_fit}
visualize_fit <- function(rep_data, params, r_R, r_P, N_R0 = 40, N_P0 = 80) {
  full_params <- c(r_R = r_R, r_P = r_P, params)
  times <- seq(0, max(rep_data$day), length.out = 200)
  sim <- simulate_lv(full_params, times = times, N_R0 = N_R0, N_P0 = N_P0)

  sim_long <- sim %>%
    pivot_longer(cols = c(N_R, N_P), names_to = "species", values_to = "density")

  obs_long <- rep_data %>%
    pivot_longer(cols = c(N_R, N_P), names_to = "species", values_to = "density")

  ggplot() +
    geom_point(data = obs_long,
               aes(x = day, y = density, color = species), size = 2) +
    geom_line(data = sim_long,
              aes(x = time, y = density, color = species)) +
    scale_y_continuous(trans = "log10") +
    labs(x = "Day", y = "Density (per mL, log scale)",
         title = paste0("LV fit: ", rep_data$treatID[1],
                       ", rep ", rep_data$repNum[1])) +
    theme_minimal()
}

visualize_fit(test_rep, est_params, r_R_test, r_P_test)
```

# Batch fit all competition replicates

```{r batch_fit}
fit_lv_one <- function(rep_data) {
  evol <- as.character(rep_data$evolvedTemp[1])
  curr <- as.character(rep_data$currentTemp[1])

  r_R <- r_rotifer_lookup %>%
    filter(evolvedTemp == evol, currentTemp == curr) %>%
    pull(r_R)
  r_P <- r_protist_lookup %>%
    filter(currentTemp == curr) %>%
    pull(r_P)
  a_RR_i <- a_rotifer_lookup %>%
    filter(evolvedTemp == evol, currentTemp == curr) %>%
    pull(a_RR_init)
  a_PP_i <- a_protist_lookup %>%
    filter(currentTemp == curr) %>%
    pull(a_PP_init)

  log_init <- log(c(a_RR_i, a_RR_i * 0.5, a_PP_i, a_PP_i * 0.5))

  fit <- tryCatch(
    optim(
      par = log_init, fn = nll_lv,
      data = rep_data, r_R = r_R, r_P = r_P,
      N_R0 = 40, N_P0 = 80,
      method = "Nelder-Mead",
      control = list(maxit = 5000, reltol = 1e-8)
    ),
    error = function(e) NULL
  )

  if (is.null(fit) || fit$convergence != 0) {
    return(tibble(
      treatID = rep_data$treatID[1], repNum = rep_data$repNum[1],
      evolvedTemp = evol, currentTemp = curr,
      a_RR = NA, a_RP = NA, a_PP = NA, a_PR = NA,
      ssq = NA, converged = FALSE
    ))
  }

  est <- exp(fit$par)
  tibble(
    treatID = rep_data$treatID[1], repNum = rep_data$repNum[1],
    evolvedTemp = evol, currentTemp = curr,
    a_RR = est[1], a_RP = est[2], a_PP = est[3], a_PR = est[4],
    ssq = fit$value, converged = TRUE
  )
}

# Split into list of replicates and fit each
rep_list <- comp_wide %>%
  group_by(treatID, repNum) %>%
  group_split()

fit_results <- map_dfr(rep_list, fit_lv_one)

# Convergence summary
cat("Total replicates:", nrow(fit_results), "\n")
cat("Converged:", sum(fit_results$converged), "\n")
cat("Failed to converge:", sum(!fit_results$converged), "\n\n")

# Distribution of parameter estimates
fit_results %>%
  filter(converged) %>%
  summarise(across(c(a_RR, a_RP, a_PP, a_PR),
                   list(median = ~ median(.x, na.rm = TRUE),
                        min = ~ min(.x, na.rm = TRUE),
                        max = ~ max(.x, na.rm = TRUE))))
```

# Diagnostics

Visualize a sample of fits across treatments to assess fit quality.

```{r diagnostic_plots}
sample_fits <- fit_results %>%
  filter(converged) %>%
  group_by(treatID) %>%
  slice_head(n = 2) %>%
  ungroup()

for (i in seq_len(nrow(sample_fits))) {
  row <- sample_fits[i, ]
  rep_data <- comp_wide %>%
    filter(treatID == row$treatID, repNum == row$repNum)
  r_R <- r_rotifer_lookup %>%
    filter(evolvedTemp == row$evolvedTemp, currentTemp == row$currentTemp) %>%
    pull(r_R)
  r_P <- r_protist_lookup %>%
    filter(currentTemp == row$currentTemp) %>%
    pull(r_P)
  params <- c(a_RR = row$a_RR, a_RP = row$a_RP,
              a_PP = row$a_PP, a_PR = row$a_PR)
  print(visualize_fit(rep_data, params, r_R, r_P))
}
```

# Compare to single-species logistic fits

The LV-derived `a_RR` and `a_PP` should be in the ballpark of the single-species
monoculture alphas if the model is well-identified. Large discrepancies
suggest either competition is meaningfully altering self-suppression dynamics
or the LV fit is absorbing variation into the wrong terms.

```{r compare_to_logistic}
comparison <- fit_results %>%
  filter(converged) %>%
  left_join(a_rotifer_lookup, by = c("evolvedTemp", "currentTemp")) %>%
  left_join(a_protist_lookup, by = "currentTemp") %>%
  group_by(evolvedTemp, currentTemp) %>%
  summarise(
    a_RR_LV_median = median(a_RR),
    a_RR_mono = first(a_RR_init),
    a_PP_LV_median = median(a_PP),
    a_PP_mono = first(a_PP_init),
    .groups = "drop"
  )
print(comparison)
```

# Treatment-level analysis

ANOVA-style models on the four LV parameters across evolved temperature and
current temperature.

```{r treatment_analysis}
fit_clean <- fit_results %>%
  filter(converged) %>%
  mutate(
    evolvedTemp = factor(evolvedTemp),
    currentTemp = factor(currentTemp),
    log_a_RR = log(a_RR),
    log_a_RP = log(a_RP),
    log_a_PP = log(a_PP),
    log_a_PR = log(a_PR)
  )

# Per-capita effect of rotifers on protists
cat("\n=== a_PR (rotifer effect on protist) ===\n")
m_PR <- lm(log_a_PR ~ evolvedTemp * currentTemp, data = fit_clean)
print(car::Anova(m_PR, type = "III"))

# Per-capita effect of protists on rotifers
cat("\n=== a_RP (protist effect on rotifer) ===\n")
m_RP <- lm(log_a_RP ~ evolvedTemp * currentTemp, data = fit_clean)
print(car::Anova(m_RP, type = "III"))

# Rotifer self-suppression
cat("\n=== a_RR (rotifer self-suppression) ===\n")
m_RR <- lm(log_a_RR ~ evolvedTemp * currentTemp, data = fit_clean)
print(car::Anova(m_RR, type = "III"))

# Protist self-suppression
cat("\n=== a_PP (protist self-suppression) ===\n")
m_PP <- lm(log_a_PP ~ evolvedTemp * currentTemp, data = fit_clean)
print(car::Anova(m_PP, type = "III"))
```

# Coexistence analysis

For each treatment, check the standard coexistence condition: stable
coexistence in two-species Lotka-Volterra requires intraspecific competition
to exceed interspecific competition for both species. This translates to:
$\alpha_{RR}/\alpha_{PR} > \alpha_{RP}/\alpha_{PP}$ (rotifer can invade
protist) and the reverse.

```{r coexistence}
coex <- fit_clean %>%
  mutate(
    rotifer_invades = a_RR / a_PR,   # > 1 means rotifer self-limits more
                                     # than it limits protists
    protist_invades = a_PP / a_RP,   # > 1 means protist self-limits more
                                     # than it limits rotifers
    coexistence = rotifer_invades > 1 & protist_invades > 1
  ) %>%
  group_by(evolvedTemp, currentTemp) %>%
  summarise(
    n_replicates = n(),
    median_rotifer_invades = median(rotifer_invades),
    median_protist_invades = median(protist_invades),
    n_predicting_coexistence = sum(coexistence),
    .groups = "drop"
  )
print(coex)
```

# Save results

```{r save_results}
saveRDS(fit_results, "lv_fit_results.rds")
saveRDS(fit_clean, "lv_fit_clean.rds")
```

# Notes and caveats

1. **Fixed r values**: The model fixes intrinsic growth rates at monoculture
   estimates. If rotifers or protists have systematically different growth
   rates under competition for reasons beyond resource depletion (e.g.,
   chemical interference, behavioral changes), those effects will be
   absorbed into the alpha estimates, biasing them.

2. **Identifiability**: Two-species LV with four free alphas has known
   identifiability issues, especially when one species' density is much
   higher than the other or when the dynamics don't show strong feedback.
   Watch for cases where standard errors of estimates are very wide.

3. **Initial conditions**: Assumed at 40 rotifers/mL and 80 protists/mL.
   If actual starting densities varied across replicates, this will
   introduce error.

4. **Time units**: All rates and alphas are now on per-day units. If the
   original fitting was per-hour and the data is per-day, this needs to be
   consistent throughout.

5. **Convergence**: Replicates that fail to converge are dropped. If many
   fail, consider trying different initial values or a different optimizer
   (e.g., L-BFGS-B with explicit bounds).
