---
title: "mendelson_thesis"
output: 
  html_document: 
    toc: true
---

# Data import

```{r}

library(tidyverse)

exp1_growthcurves <- read.csv("../data/exp_1_growthcurves.csv")
exp1_bodysize     <- read.csv("../data/exp_1_size.csv")

exp2_growthcurves <- read.csv("../data/exp_2_growthcurves.csv")
exp2_bodysize    <- read.csv("../data/exp_2_size.csv")

```

# Experiment 1

*Diversity of response in temperature*

### Data structuring

```{r}

exp1_growthcurves <- exp1_growthcurves %>%
  mutate(Day = Day - 3,
         Count1 = Count + 1,
         across(c(Site, Leaf, Clone, Treatment, ID), as.factor),
         across(c(Day, Count), as.integer))

clone_key <- c(
  "1" = 1,  "2" = 2,  "17" = 3,  "18" = 4,  "19" = 5,
  "72" = 6, "73" = 7, "74" = 8, "39" = 9,  "40" = 10,
  "42" = 11,"75" = 12,"26" = 13,"25" = 14,"35" = 15,
  "36" = 16,"37" = 17
  )

exp1_bodysize <- exp1_bodysize %>%
  separate(base, into = c("unk", "culture", "image", "object"), 
           sep = "_", convert = TRUE) %>%
  separate(culture, into = c("siteLeaf", "clone", "replicate"), 
           sep = "-", convert = TRUE) %>%
  mutate(site = as.integer(case_when(str_sub(siteLeaf, 1, 1) == "c" ~ 1,
                                     str_sub(siteLeaf, 1, 1) == "p" ~ 2)),
         leaf = str_sub(siteLeaf, start = 2),
         leaf = as.integer(case_when(leaf == "23" ~ 1,
                                     leaf == "7"  ~ 2,
                                     leaf == "36" & site == "1" ~ 3,
                                     leaf == "36" & site == "2" ~ 4,
                                     leaf == "24" ~ 5,
                                     leaf == "3"  ~ 6)),
         clone = as.integer(unname(clone_key[as.character(clone)])),
         temperature = as.factor(case_when(replicate %in% 
                                             c(1, 2, 3) ~ "25C",
                                           replicate %in% 
                                             c(4, 5, 6) ~ "30C")),
         Tr = ifelse(temperature == "30C", 0.5, -0.5),
        across(c(temperature, site, leaf, clone, replicate), factor),
        area_um = area_px * (200 / 139)^2) %>%
  filter_out(clone=="6")

```

### Growth Models

```{r}

library(growthrates)

exp1_growth_models <- all_splines(
                  Count1 ~ Day | Treatment + Site + Leaf + Clone + ID,
                  data = exp1_growthcurves,
                  spar = 0.5
                  )

# helper function to visualize model fits
plot_exp1_models <- function(models) {
  coefs <- as.data.frame(coef(models))
  model_ids <- c(1:length(coefs$y0))
  
  columns <- floor(sqrt(length(model_ids)))
  rows <- columns + 1
  par(mfrow = c(columns, rows), mar = c(1, 1, 1, 1))
  
  for (i in seq_along(model_ids)) {
    plot(models[[i]])
    
    this_r <- round(coefs$mumax[i], 6)
    this_model <- rownames(coefs)[i]
    
    mtext(this_model, side = 3, adj = 0.1, line = -1, cex = 0.5)
    mtext(paste("r:", this_r), side = 3, adj = 0.1, line = -2, cex = 0.5)
  }
}

get_t_mumax_i <- function(i, models) {
  models[[i]]@xy[1]
}

get_max_abundance_i <- function(i, models) {
  max(models[[i]]@obs$y, na.rm = TRUE)
}

exp1_growth_summary <- results(exp1_growth_models) %>%
  mutate(
    across(c(Treatment, Site, Leaf, Clone), factor),

    Tr = ifelse(Treatment == "30C", 0.5, -0.5),

    t_mumax = vapply(
      seq_len(n()),
      get_t_mumax_i,
      numeric(1),
      models = exp1_growth_models
    ),

    tpeak_cat = cut(
      t_mumax,
      breaks = c(-Inf, 3, 6, 9, Inf),
      labels = c("0-3", "3-6", "6-9", ">9"),
      ordered_result = TRUE
    ),

    max_abundance = vapply(
      seq_len(n()),
      get_max_abundance_i,
      numeric(1),
      models = exp1_growth_models
    )
  )


```



### Exp1 Analysis

```{r}

library(car)
library(moments)
library(lme4)
library(lmerTest)


### Maximum growth rate:

# raw data for variance components
exp1_lmm_r <- lmer(mumax ~ 
                       Tr + 
                       (1 | Site) +
                       (1 | Site:Leaf) + 
                       (0+Tr | Site:Leaf) + 
                       (1 | Site:Leaf:Clone) + 
                       (0+Tr | Site:Leaf:Clone), 
                       data = exp1_growth_summary
                     )

summary(exp1_lmm_r)

hist(exp1_growth_summary$mumax)
hist(log(exp1_growth_summary$mumax))

# log-transformed normal data for fixed effect
exp1_lmm_log_r <- lme4::lmer(log(mumax) ~ 
                         Tr + 
                         (1 | Site:Leaf) + 
                         (0+Tr | Site:Leaf) + 
                         (1 | Site:Leaf:Clone) + 
                         (0+Tr | Site:Leaf:Clone), 
                       data = exp1_growth_summary
                     )

plot(exp1_lmm_log_r)
qqp(resid(exp1_lmm_log_r), "norm")
skewness(log(exp1_growth_summary$mumax))
kurtosis(log(exp1_growth_summary$mumax))



exp1_lmm_log_r_2 <- lme4::lmer(log(mumax) ~ 
                         Tr +
                         (1 | Site:Leaf:Clone) + 
                         (0+Tr | Site:Leaf:Clone), 
                       data = exp1_growth_summary
                     )

Anova(exp1_lmm_log_r_2, type = "III", test.statistic = "F")

### Body size:

# exp1_lmm_size <- lmer(area_px ~ 
#                        Tr + 
#                        (1 | site) + 
#                        (1 | site:leaf) + 
#                        (1 | site:leaf:clone), 
#                     data = exp1_bodysize)
# 
# summary(exp1_lmm_size)

# exp1_lmm_size_no_site <- lmer(area_px ~ 
#                        Tr + 
#                        (1 | site:leaf) + 
#                        (1 | site:leaf:clone), 
#                     data = exp1_bodysize)
# anova(exp1_lmm_size, exp1_lmm_size_no_site)

exp1_lmm_size_no_leaf <- lmer(area_px ~ 
                       Tr + 
                       (1 | site) +
                       (1 | site:leaf:clone), 
                    data = exp1_bodysize)
anova(exp1_lmm_size, exp1_lmm_size_no_leaf)

hist(exp1_bodysize$area_um)
hist(sqrt(exp1_bodysize$area_um))

# square root transformed data for fixed effect
exp1_lmm_sqrt_size <- lmer(sqrt(area_um) ~ 
                       Tr + 
                       (1 | site) + 
                       (1 | site:leaf) + 
                       (1 | site:leaf:clone), 
                    data = exp1_bodysize)

plot(exp1_lmm_sqrt_size)
qqp(resid(exp1_lmm_sqrt_size), "norm")
qqp(resid(exp1_lmm_size), "norm")
skewness(sqrt(exp1_bodysize$area_um))
kurtosis(sqrt(exp1_bodysize$area_um))

Anova(exp1_lmm_sqrt_size, type = "III")

```


### Fig1

#### Code

```{r}

fig1a_data <- exp1_growth_summary %>%
  group_by(Treatment, Site, Leaf, Clone) %>%
  mutate(Treatment= recode_factor(Treatment, "25C" = "25°C", "30C" = "30°C")) %>%
  summarize(mean_mu = mean(mumax), se_mu = sd(mumax)/sqrt(length(mumax)))

fig1a <- ggplot(fig1a_data, aes(x = Treatment, y = mean_mu, color = Leaf, group = Clone)) +
            geom_point(stat = "identity", position = position_dodge(width = 0.1)) +
            geom_errorbar(stat = "identity", position = position_dodge(width = 0.1),
                          aes(ymin = mean_mu - se_mu, ymax = mean_mu + se_mu), 
                          width = 0.2, linewidth = 0.25) +
            geom_line(stat = "identity", position = position_dodge(width = 0.1)) +
            facet_wrap(~Leaf) +
            labs(x = "Temperature", y = expression("Maximum growth rate, "~italic("μ")[max]~"d "^-1)) +
            theme_minimal() +
            theme(
              panel.grid = element_blank(),
              panel.border = element_rect(linewidth = 0.5),
              strip.text = element_blank()
              )


fig1b_data <- exp1_bodysize %>%
  group_by(temperature, site, leaf, clone) %>%
  mutate(temperature = recode_factor(temperature , "25C" = "25°C", "30C" = "30°C")) %>%
  summarize(mean_size = mean(area_um), se_size = sd(area_um)/sqrt(length(area_um)))

fig1b <- ggplot(fig1b_data, aes(x = temperature, y = mean_size, color = leaf, group = clone)) +
            geom_point(stat = "identity", position = position_dodge(width = 0.1)) +
            geom_errorbar(stat = "identity", position = position_dodge(width = 0.1),
                          aes(ymin = mean_size - se_size, ymax = mean_size + se_size), 
                          width = 0.2, linewidth = 0.25) +
            geom_line(stat = "identity", position = position_dodge(width = 0.1)) +
            facet_wrap(~leaf) +
            labs(x = "Temperature", y = expression("Body size, μm "^2)) +
            theme_minimal() +
            theme(
              panel.grid = element_blank(),
              panel.border = element_rect(linewidth = 0.5),
              strip.text = element_blank()
              )

```

#### Export

```{r fig1 export}

fig1_names <- c("fig1a", "fig1b")
fig1 <- mget(fig1_names)

iwalk(fig1, ~ ggsave(
  filename = paste0(.y, ".pdf"),
  plot     = .x,
  width    = 6,
  height   = 4,
  units    = "in",
  path     = "../output/manuscriptFigures"
))

```

# Experiment 2

*Indirect effects of evolutionary history of temperature on competition in contemporary temperature environments*

### Data Structuring

```{r exp2data}
exp2_growthcurves <- exp2_growthcurves %>%
  mutate(hour = as.integer(day * 24),
         across(c(evolvedTemp, currentTemp, competition, repNum), factor))


exp2_rotifer_growth <- exp2_growthcurves %>%
  filter(species == "rotifer")
exp2_protist_growth <- exp2_growthcurves %>%
  filter(species == "protist")


exp2_bodysize <- exp2_bodysize %>%
  separate(image_name, into = c("temp_rep", "origin_filetype", "origin_filename", "image_name", "image_filetype"), 
           sep = "\\.", convert = TRUE) %>%
  separate(temp_rep, into = c("temperature", "replicate"), sep = "_", convert = TRUE) %>%
  select(temperature, replicate, area_px) %>%
  mutate(across(c(temperature, replicate), factor)) %>%
  filter(area_px > 500) %>%# removes a single erroneously misidentified object ~ 100px^2
  mutate(area_um = area_px * (200 / 139)^2) #convert to um using meas. of hemocyt. image

exp2_initial_bodysize <- exp1_bodysize %>%
  select(temperature, clone, area_um)

```

### Growth Models

```{r exp2logistic}

library(growthrates)

# helper for plotting models
plot_exp2_models <- function(models) {
  coefs <- as.data.frame(coef(models))
  model_ids <- c(1:length(coefs$y0))
  
  n <- length(model_ids)
  columns <- floor(sqrt(n))
  rows <- ceiling(n / columns)
  par(mfrow = c(columns, rows), mar = c(1, 1, 1, 1))
  
  for (i in seq_along(model_ids)) {
    plot(models[[i]])
    
    this_r <- round(coefs$r[i], 6)
    this_a <- round(coefs$alpha[i], 6)
    this_model <- rownames(coefs)[i]
    
    mtext(this_model, side = 3, adj = 0.1, line = -1, cex = 0.5)
    mtext(paste("r:", this_r), side = 3, adj = 0.1, line = -2, cex = 0.5)
    mtext(paste("a:", this_a), side = 3, adj = 0.1, line = -2, cex = 0.5)
  }
}

# user-specified model
logistic_alpha_function <- function(time, parms) {
  y0    <- parms[["y0"]]
  r     <- parms[["r"]]
  alpha <- parms[["alpha"]]
  
  y <- (r * y0) / (alpha * y0 + (r - alpha * y0) * exp(-r * time))
  
  cbind(time = time, y = y)
}

grow_logistic_alpha <- growthmodel(
  logistic_alpha_function,
  pnames = c("y0", "r", "alpha")
)

exp2_growth_models_rotifer <- all_growthmodels(
  count ~ hour | evolvedTemp + currentTemp + competition + repNum, 
    data  = exp2_rotifer_growth,
    p     = c(y0 = 5, r = 0.1, alpha = 0.0005), # initial
    upper = c(y0 = 7, r = 8, alpha = 1),
    lower = c(y0 = 3, r = 0.003, alpha = 0.0001),
    FUN   = grow_logistic_alpha
)

exp2_growth_models_protist <- all_growthmodels(
  count ~ hour | evolvedTemp + currentTemp + competition + repNum, 
    data  = exp2_protist_growth,
    p     = c(y0 = 10, r = 0.1, alpha = 0.0003), # initial
    upper = c(y0 = 15, r = 5, alpha = 0.02),
    lower = c(y0 = 5, r = 0.001, alpha = 0.000005),
    FUN   = grow_logistic_alpha
)

# model summaries

exp2_growth_summary_rotifer <- results(exp2_growth_models_rotifer) %>%
  rownames_to_column(var = "id") %>%
  separate(col = "id", 
           into = c("evolvedTemp", "currentTemp", "competition", "replicate"), 
           sep = ":") %>%
  mutate(across(c(evolvedTemp, currentTemp, competition, replicate), factor),
         r_hour = r, r_day = r * 24,
         alpha_hour = alpha, alpha_day = alpha * 24,
         K = (r_day / alpha_day) * 10, # x10 is to convert from 0.1ml to 1ml
         competition = case_when(competition == TRUE ~ "Present",
                                 competition == FALSE ~ "Absent"))

exp2_growth_summary_protist <- results(exp2_growth_models_protist) %>%
  rownames_to_column(var = "id") %>%
  separate(col = "id", 
           into = c("evolvedTemp", "currentTemp", "competition", "replicate"), 
           sep = ":") %>%
  mutate(across(c(evolvedTemp, currentTemp, competition, replicate), factor),
         compFactor = case_when(
           evolvedTemp == "25" & competition == TRUE ~ "25-evol-rotif",
           evolvedTemp == "30" & competition == TRUE ~ "30-evol-rotif",
           competition == FALSE ~ "no-comp"),
         r_hour = r, r_day = r * 24,
         alpha_hour = alpha, alpha_day = alpha * 24,
         K = (r_day / alpha_day) * 10) # x10 is to convert from 0.1ml to 1ml

```

### Linear Model Analysis

```{r exp2lm}

library(lme4)
library(lmerTest)
library(car)
library(moments)
library(emmeans)


### Rotifers: Intrinsic growth rate

hist(exp2_growth_summary_rotifer$r_day)
hist(log(exp2_growth_summary_rotifer$r_day))

exp2_lm_rotifer_log_r <- lm(log(r_day) ~ evolvedTemp * currentTemp * competition, 
                            data = exp2_growth_summary_rotifer)

plot(exp2_lm_rotifer_log_r)
qqp(exp2_lm_rotifer_log_r, "norm")
skewness(log(exp2_growth_summary_rotifer$r_day))
kurtosis(log(exp2_growth_summary_rotifer$r_day))

Anova(exp2_lm_rotifer_log_r, type = "III")

emm_rotifer_r_EvxCo <- emmeans(exp2_lm_rotifer_log_r, ~ evolvedTemp | competition)
pairs(emm_rotifer_r_EvxCo, adjust = "tukey")

emm_rotifer_r_CoxEv <- emmeans(exp2_lm_rotifer_log_r, ~ competition | evolvedTemp)
pairs(emm_rotifer_r_CoxEv, adjust = "tukey")


emm_rotifer_r_CuxCo <- emmeans(exp2_lm_rotifer_log_r, ~ currentTemp | competition)
pairs(emm_rotifer_r_CuxCo, adjust = "tukey")

emm_rotifer_r_CoxCu <- emmeans(exp2_lm_rotifer_log_r, ~ competition | currentTemp)
pairs(emm_rotifer_r_CoxCu, adjust = "tukey")


emm_rotifer_r_EvxCuxCo <- emmeans(exp2_lm_rotifer_log_r,
               ~ evolvedTemp | currentTemp * competition)

pairs(emm_rotifer_r_EvxCuxCo, adjust = "tukey")

emm_rotifer_r_FULL <- emmeans(exp2_lm_rotifer_log_r, ~ evolvedTemp * currentTemp * competition)
pairs(emm_rotifer_r_FULL, adjust = "tukey")

library(multcomp)

rotifer_r_cld <- multcomp::cld(
  emm_rotifer_r_FULL,
  adjust = "tukey",
  Letters = letters
)

# emmeans SE is 0.0772 

### Rotifers: Crowding coefficient

hist(exp2_growth_summary_rotifer$alpha_day)
qqp(exp2_growth_summary_rotifer$alpha_day, "norm")

exp2_lm_rotifer_alpha <- lm(alpha_day ~ evolvedTemp * currentTemp * competition, 
                            data = exp2_growth_summary_rotifer)

plot(exp2_lm_rotifer_alpha)
qqp(exp2_lm_rotifer_alpha, "norm")
skewness(exp2_growth_summary_rotifer$alpha_day)
kurtosis(exp2_growth_summary_rotifer$alpha_day)

Anova(exp2_lm_rotifer_alpha, type = "III")

emm_rotifer_alpha_CuxCo <- emmeans(exp2_lm_rotifer_alpha, ~ currentTemp | competition)
pairs(emm_rotifer_alpha_CuxCo, adjust = "tukey")


emm_rotifer_alpha_CuxCoxEv <- emmeans(exp2_lm_rotifer_alpha, ~ evolvedTemp | competition * currentTemp)
pairs(emm_rotifer_alpha_CuxCoxEv, adjust = "tukey")

emm_rotifer_alpha_FULL <- emmeans(exp2_lm_rotifer_alpha, ~ evolvedTemp * currentTemp * competition)
pairs(emm_rotifer_alpha_FULL, adjust = "tukey")

### Rotifers: Equilibrium

hist(exp2_growth_summary_rotifer$K)
qqp(exp2_growth_summary_rotifer$K, 'norm')

exp2_lm_rotifer_eq <- lm(K ~ evolvedTemp * currentTemp * competition,
                         data = exp2_growth_summary_rotifer)

plot(exp2_lm_rotifer_eq)
qqp(exp2_lm_rotifer_eq, "norm")
skewness(exp2_growth_summary_rotifer$alpha_day)
kurtosis(exp2_growth_summary_rotifer$alpha_day)

Anova(exp2_lm_rotifer_eq, type = "III")

### Protists: Intrinsic growth rate

hist(exp2_growth_summary_protist$r_day)
hist(log(exp2_growth_summary_protist$r_day))

exp2_lm_protist_log_r <- lm(log(r_day) ~ currentTemp * compFactor, 
                            data = exp2_growth_summary_protist)

plot(exp2_lm_protist_log_r)
qqp(exp2_lm_protist_log_r, "norm")
skewness(log(exp2_growth_summary_protist$r_day))
kurtosis(log(exp2_growth_summary_protist$r_day))

Anova(exp2_lm_protist_log_r, type = "III")

emm_protist_r_CuxCo <- emmeans(exp2_lm_protist_log_r, ~ currentTemp | compFactor)
pairs(emm_protist_r_CuxCo, adjust = "tukey")

emm_protist_r_CoxCu <- emmeans(exp2_lm_protist_log_r, ~ compFactor | currentTemp)
pairs(emm_protist_r_CoxCu, adjust = "tukey")

### Protists: Crowding coefficient

hist(exp2_growth_summary_protist$alpha_day)
hist(log(exp2_growth_summary_protist$alpha_day))

exp2_lm_protist_log_alpha <- lm(log(alpha_day) ~ currentTemp * compFactor, 
                            data = exp2_growth_summary_protist)

plot(exp2_lm_protist_log_alpha)
qqp(exp2_lm_protist_log_alpha, "norm")
skewness(log(exp2_growth_summary_protist$alpha_day))
kurtosis(log(exp2_growth_summary_protist$alpha_day))

Anova(exp2_lm_protist_log_alpha, type = "III")

emm_protist_alpha_CuxCo <- emmeans(exp2_lm_protist_log_alpha, ~ currentTemp | compFactor)
pairs(emm_protist_alpha_CuxCo, adjust = "tukey")

emm_protist_alpha_CoxCu <- emmeans(exp2_lm_protist_log_alpha, ~ compFactor | currentTemp)
pairs(emm_protist_alpha_CoxCu, adjust = "tukey")


### Protists: Equilibrium

hist(exp2_growth_summary_protist$K)
qqp(exp2_growth_summary_protist$K, 'norm')

exp2_lm_protist_eq <- lm(K ~ currentTemp * compFactor,
                         data = exp2_growth_summary_protist)

plot(exp2_lm_protist_eq)
qqp(exp2_lm_protist_eq, "norm")
skewness(exp2_growth_summary_protist$alpha_day)
kurtosis(exp2_growth_summary_protist$alpha_day)

Anova(exp2_lm_protist_eq, type = "III")

emm_protist_eq_CuxCo <- emmeans(exp2_lm_protist_eq, ~ currentTemp | compFactor)
emm_protist_eq_CoxCu <- emmeans(exp2_lm_protist_eq, ~ compFactor | currentTemp)

pairs(emm_protist_eq_CuxCo)
pairs(emm_protist_eq_CoxCu)

### Rotifer size

hist(exp2_bodysize$area_um)
qqp(exp2_bodysize$area_um, 'norm')

hist(log(exp2_bodysize$area_um))
qqp(log(exp2_bodysize$area_um), 'norm')

exp2_lmm_log_size <- lmerTest::lmer(log(area_um) ~ temperature + 
                            (1|temperature:replicate), data = exp2_bodysize)

exp2_lm_log_size <- lm(log(area_um) ~ temperature, data = exp2_bodysize)

anova(exp2_lmm_log_size,exp2_lm_log_size)

anova(exp2_lm_log_size)
Anova(exp2_lm_log_size, type = "III")

plot(exp2_lmm_log_size)
qqp(resid(exp2_lmm_log_size))
hist(resid(exp2_lmm_log_size), breaks = 30)

summary(exp2_lmm_log_size)
Anova(exp2_lmm_log_size, type = "III")

### Size comparison with initial

initial_equal_weighted_size <- exp2_initial_bodysize %>%
  group_by(clone) %>%
  summarize(clone_mean = mean(area_um)) %>%
  summarize(mean = mean(clone_mean),
            se = sd(clone_mean) / sqrt(n()),
            n = n())

t.test(exp2_bodysize %>% filter(temperature == 25) %>% pull(area_um),
       initial_size,
       alternative = "two.sided")

t.test(exp2_bodysize %>% filter(temperature == 30) %>% pull(area_um),
       initial_size,
       alternative = "two.sided")

t.test(area_um ~ temperature, data = exp2_bodysize)

### Growth rate comparison with initial



```



### Table Exports

```{r tables}

table2.x <- results(exp2_growth_models_rotifer) %>%
  summarize(meanr2 = mean(r2),
            sdr2 = sd(r2))

table2.y <- results(exp2_growth_models_protist) %>%
  summarize(meanr2 = mean(r2),
            sdr2 = sd(r2))

table2.1 <- Anova(exp2_lm_rotifer_log_r, type = "III")

stargazer(
table2.1,
type = "html",
out = "table2.1.html",
summary = FALSE,
rownames = TRUE,
title = "Table 2.1"
)


table2.2 <- as.data.frame(pairs(emm_rotifer_r_EvxCuxCo, adjust = "tukey"))

stargazer(
table2.2,
type = "html",
out = "table2.2.html",
summary = FALSE,
rownames = TRUE,
title = "Table 2.2"
)


table2.3 <- Anova(exp2_lm_rotifer_alpha, type = "III")

stargazer(
table2.3,
type = "html",
out = "table2.3.html",
summary = FALSE,
rownames = TRUE,
title = "Table 2.3"
)


table2.4 <- as.data.frame(pairs(emm_rotifer_alpha_CuxCo, adjust = "tukey"))

stargazer(
table2.4,
type = "html",
out = "table2.4.html",
summary = FALSE,
rownames = TRUE,
title = "Table 2.4"
)

```

### Generations

```{r gen}

t_evol <- 135 # days spent in constant temperature

g_rotifers <- exp2_growth_summary_rotifer %>% 
  filter(competition == "Absent", evolvedTemp == currentTemp) %>%
  group_by(evolvedTemp, currentTemp) %>%
  summarize(g = log(2)/mean(r_day),
            t_evol = t_evol,
            gs_evol = as.integer(t_evol / g))

g_protists <- exp2_growth_summary_protist %>%
  filter(compFactor == "no-comp") %>%
  group_by(evolvedTemp, currentTemp) %>%
  summarize(g = log(2)/mean(r_day))


```

### Fig 2

#### Code

```{r fig2 code}

fig2a_data <- as.data.frame(emmeans(exp2_lm_rotifer_log_r, ~ evolvedTemp * currentTemp * competition)) %>%
  mutate(currentTemp = recode_factor(currentTemp, "25" = "Contemporary 25°C", "30" = "Contemporary 30°C"),
         evolvedTemp = recode_factor(evolvedTemp, "25" = "25°C", "30" = "30°C"))

fig2a <- ggplot(fig2a_data, aes(x = competition, y = emmean, color = evolvedTemp, shape = evolvedTemp, group = evolvedTemp)) +
            geom_point(stat = "identity", position = position_dodge(width = 0.1)) +
            geom_errorbar(stat = "identity", position = position_dodge(width = 0.1),
                          aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.15) +
            geom_line(stat = "identity", position = position_dodge(width = 0.1)) +
            facet_wrap(~currentTemp) +
            labs(x = "Competition", y = expression("Rotifer growth rate, ln("*italic(r)*")"),
                 color = "History of Temp") +
            theme_minimal() +
            scale_color_manual(values = c("#1BB6AF", "#E9A17C")) +
            guides(shape = "none") +
            theme(
              panel.grid = element_blank(),
              panel.border = element_rect(linewidth = 0.5),
              legend.position = "bottom"
              )

fig2at <- ggplot(fig2a_data, aes(x = competition, y = emmean, color = evolvedTemp, group = evolvedTemp)) +
            geom_rect(aes(fill = currentTemp),
              xmin = -Inf, xmax = Inf,
              ymin = -Inf, ymax = Inf,
              inherit.aes = FALSE,
              alpha = 0.05) +
            scale_fill_manual(values = c("#1BB6AF", "#E9A17C")) +
            geom_point(stat = "identity", position = position_dodge(width = 0.1)) +
            geom_errorbar(stat = "identity", position = position_dodge(width = 0.1),
                          aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.15) +
            geom_line(stat = "identity", position = position_dodge(width = 0.1)) +
            facet_wrap(~currentTemp) +
            labs(x = "Competition", y = "ln(Intrinsic growth rate)") +
            coord_cartesian(ylim = c(-1.2, -0.2)) +
            theme_minimal() +
            scale_color_manual(values = c("#1BB6AF", "#E9A17C")) +
            theme(
              panel.grid = element_blank(),
              panel.border = element_rect(linewidth = 0.5),
              legend.position = "none",
              panel.background = element_rect(fill = "transparent", color = NA)
              )


fig2a_combo_data <- as.data.frame(emmeans(exp2_lm_rotifer_log_r, ~ currentTemp * competition))

fig2a_combo <- ggplot(fig2a_combo_data, aes(x = competition, y = emmean, group = currentTemp)) +
            geom_rect(aes(fill = currentTemp),
                        xmin = -Inf, xmax = Inf,
                        ymin = -Inf, ymax = Inf,
                        inherit.aes = FALSE,
                        alpha = 0.1) +
            scale_fill_manual(values = c("#1BB6AF", "#E9A17C")) +
            geom_point(stat = "identity", position = position_dodge(width = 0.1)) +
            geom_errorbar(stat = "identity", position = position_dodge(width = 0.1),
                          aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.15) +
            geom_line(stat = "identity", position = position_dodge(width = 0.1)) +
            facet_wrap(~currentTemp) +
            labs(x = "Competition", y = "ln(Intrinsic growth rate)") +
            coord_cartesian(ylim = c(-1.2, -0.2)) +
            theme_minimal() +
            scale_color_manual(values = c("#1BB6AF", "#E9A17C")) +
            theme(
              panel.grid = element_blank(),
              panel.border = element_rect(linewidth = 0.5),
              legend.position = "none",
              panel.background = element_rect(fill = "transparent", color = NA)
              )


fig2b_data <- as.data.frame(emmeans(exp2_lm_rotifer_alpha, ~ evolvedTemp * currentTemp * competition)) %>%
  mutate(currentTemp = recode_factor(currentTemp, "25" = "Contemporary 25°C", "30" = "Contemporary 30°C"),
           evolvedTemp = recode_factor(evolvedTemp, "25" = "25°C", "30" = "30°C"))

fig2b <- ggplot(fig2b_data, aes(x = competition, y = emmean, color = evolvedTemp, shape = evolvedTemp, group = evolvedTemp)) +
            geom_point(stat = "identity", position = position_dodge(width = 0.1)) +
            geom_errorbar(stat = "identity", position = position_dodge(width = 0.1),
                          aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.15) +
            geom_line(stat = "identity", position = position_dodge(width = 0.1)) +
            facet_wrap(~currentTemp) +
            labs(x = "Competition", y = "Rotifer crowding coefficient, ln(α)",
                 color = "History of Temp") +
            theme_minimal() +
            guides(shape = "none") +
            scale_color_manual(values = c("#1BB6AF", "#E9A17C")) +
            theme(
              panel.grid = element_blank(),
              panel.border = element_rect(linewidth = 0.5),
              legend.position = "bottom"
              )


fig2c_data <- as.data.frame(emmeans(exp2_lm_protist_log_r, ~ currentTemp * compFactor)) %>%
  mutate(currentTemp = recode_factor(currentTemp, "25" = "25°C", "30" = "30°C"),
         compFactor = recode_factor(compFactor, "25-evol-rotif" = "Rotifers evo 25°C", "30-evol-rotif" = "Rotifers evo 30°C",
                                     "no-comp" = "Protists only"))

fig2c <- ggplot(fig2c_data, aes(x = currentTemp, y = emmean, color = compFactor, shape = compFactor, group = compFactor)) +
            geom_point(stat = "identity", position = position_dodge(width = 0.1)) +
            geom_errorbar(stat = "identity", position = position_dodge(width = 0.1),
                          aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.15) +
            geom_line(stat = "identity", position = position_dodge(width = 0.1)) +
            labs(x = "Contemporary Temperature", y = expression("Protist growth rate, ln("*italic(r)*")"),
                 color = "Competition with:", shape = "Competition with:") +
            theme_minimal() +
            scale_color_manual(values = c("#1BB6AF", "#E9A17C", "lightgreen")) +
            theme(
              panel.grid = element_blank(),
              panel.border = element_rect(linewidth = 0.5),
              legend.position = "bottom"
              )


fig2d_data <- as.data.frame(emmeans(exp2_lm_protist_log_alpha, ~ currentTemp * compFactor)) %>%
    mutate(currentTemp = recode_factor(currentTemp, "25" = "25°C", "30" = "30°C"),
         compFactor = recode_factor(compFactor, "25-evol-rotif" = "Rotifers evo 25°C", 
                                    "30-evol-rotif" = "Rotifers evo 30°C", "no-comp" = "Protists only"))

fig2d <- ggplot(fig2d_data, aes(x = currentTemp, y = emmean, color = compFactor, shape = compFactor, group = compFactor)) +
            geom_point(stat = "identity", position = position_dodge(width = 0.1)) +
            geom_errorbar(stat = "identity", position = position_dodge(width = 0.1),
                          aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.15) +
            geom_line(stat = "identity", position = position_dodge(width = 0.1)) +
            labs(x = "Contemporary Temperature", y = "Protist crowding coefficient, ln(α)",
                 shape = "Competition with:", color = "Competition with:") +
            theme_minimal() +
            scale_color_manual(values = c("#1BB6AF", "#E9A17C", "lightgreen")) +
            theme(
              panel.grid = element_blank(),
              panel.border = element_rect(linewidth = 0.5),
              legend.position = "bottom"
              )

fig2e_data <-as.data.frame(emmeans(exp2_lm_rotifer_eq, ~ evolvedTemp * currentTemp * competition)) %>%
  mutate(currentTemp = recode_factor(currentTemp, "25" = "Contemporary 25°C", "30" = "Contemporary 30°C"),
           evolvedTemp = recode_factor(evolvedTemp, "25" = "25°C", "30" = "30°C"))

fig2e <- ggplot(fig2e_data, aes(x = competition, y = emmean, color = evolvedTemp, shape = evolvedTemp, group = evolvedTemp)) +
            geom_point(stat = "identity", position = position_dodge(width = 0.1)) +
            geom_errorbar(stat = "identity", position = position_dodge(width = 0.1),
                          aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.15) +
            geom_line(stat = "identity", position = position_dodge(width = 0.1)) +
            facet_wrap(~currentTemp) +
            labs(x = "Competition", y = expression("Rotifer realized equilibrium, "*italic(hat(N))*" ml "^-1),
                 color = "History of Temp") +
            theme_minimal() +
            guides(shape = "none") +
            scale_color_manual(values = c("#1BB6AF", "#E9A17C")) +
            theme(
              panel.grid = element_blank(),
              panel.border = element_rect(linewidth = 0.5),
              legend.position = "bottom"
              )

fig2f_data <-as.data.frame(emmeans(exp2_lm_protist_eq, ~ currentTemp * compFactor)) %>%
  mutate(currentTemp = recode_factor(currentTemp, "25" = "Contemporary 25°C", "30" = "Contemporary 30°C"),
         compFactor = recode_factor(compFactor, "25-evol-rotif" = "Rotifers evo 25°C", 
                                    "30-evol-rotif" = "Rotifers evo 30°C", "no-comp" = "Protists only"))

fig2f <- ggplot(fig2f_data, aes(x = currentTemp, y = emmean, color = compFactor, shape = compFactor, group = compFactor)) +
            geom_point(stat = "identity", position = position_dodge(width = 0.1)) +
            geom_errorbar(stat = "identity", position = position_dodge(width = 0.1),
                          aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.15) +
            geom_line(stat = "identity", position = position_dodge(width = 0.1)) +
            labs(x = "Contemporary Temperature", y = expression("Protist realized equilibrium, "*italic(hat(N))*" ml "^-1),
                 color = "Competition with:") +
            theme_minimal() +
            guides(shape = "none") +
            scale_color_manual(values = c("#1BB6AF", "#E9A17C", "lightgreen")) +
            theme(
              panel.grid = element_blank(),
              panel.border = element_rect(linewidth = 0.5),
              legend.position = "bottom"
              )


```

#### Export

```{r fig2 export}

fig2_names <- c("fig2e", "fig2f")
fig2 <- mget(fig2_names)

iwalk(fig2, ~ ggsave(
  filename = paste0(.y, ".tiff"),
  plot     = .x,
  width    = 6,
  height   = 4,
  units    = "in",
  path     = "../output/manuscriptFigures"
))

```

### Random testing

```{r}

library(dplyr)
library(tidyr)

clone_means <- growth_summary %>%
  group_by(Site, Leaf, Clone, Treatment) %>%
  summarise(
    mu_max = mean(mu_max, na.rm = TRUE),
    t_mumax = mean(t_mumax, na.rm = TRUE),
    .groups = "drop"
  )

delta_df <- clone_means %>%
  pivot_wider(
    names_from = Treatment,
    values_from = c(mu_max, t_mumax)
  ) %>%
  mutate(
    delta_mu = mu_max_30C - mu_max_25C,
    delta_t  = t_mumax_30C - t_mumax_25C
  )

delta_df
  )

cor.test(delta_df$delta_t, delta_df$delta_mu)

lm_delta <- lm(delta_mu ~ delta_t, data = delta_df)
summary(lm_delta)

library(ggplot2)

ggplot(delta_df, aes(x = delta_t, y = delta_mu)) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = expression(Delta ~ t[mu[max]]),
    y = expression(Delta ~ mu[max])
  )

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

# --------------------------------------------------
# 1. plotting data
# --------------------------------------------------
library(dplyr)

plot_df <- growth_summary %>%
  mutate(
    leaf_id = paste(Site, Leaf, sep = ":"),
    clone_id = paste(Site, Leaf, Clone, sep = ":")
  ) %>%
  group_by(leaf_id) %>%
  mutate(
    clone_shape = factor(dense_rank(Clone))
  ) %>%
  ungroup()

clone_means <- plot_df %>%
  group_by(Site, Leaf, Clone, leaf_id, clone_id, clone_shape, Treatment) %>%
  summarise(
    mu_max = mean(mu_max, na.rm = TRUE),
    t_mumax = mean(t_mumax, na.rm = TRUE),
    .groups = "drop"
  )

# --------------------------------------------------
# 3. clone-level deltas for bottom panel
# --------------------------------------------------
delta_df <- clone_means %>%
  select(Site, Leaf, Clone, leaf_id, clone_id, Treatment, mu_max, t_mumax) %>%
  pivot_wider(
    names_from = Treatment,
    values_from = c(mu_max, t_mumax)
  ) %>%
  mutate(
    delta_mu = mu_max_30C - mu_max_25C,
    delta_t  = t_mumax_30C - t_mumax_25C
  )

# --------------------------------------------------
# 4. top panel
#    - raw replicate points
#    - clone means connected across treatments
#    - clone-colored within each facet
# --------------------------------------------------
p1 <- ggplot() +
  geom_point(
    data = plot_df,
    aes(x = t_mumax, y = mu_max,
        color = Treatment,
        shape = clone_shape),
    size = 1.6,
    alpha = 0.4,
    position = position_jitter(width = 0.12, height = 0)
  ) +
  geom_line(
    data = clone_means,
    aes(x = t_mumax, y = mu_max,
        group = clone_id),
    linewidth = 0.6,
    alpha = 0.8,
    color = "gray40"
  ) +
  geom_point(
    data = clone_means,
    aes(x = t_mumax, y = mu_max,
        color = Treatment,
        shape = clone_shape),
    size = 3
  ) +
  facet_wrap(~ leaf_id, ncol = 3) +
  labs(
    x = expression(t[mu[max]]),
    y = expression(mu[max]),
    color = "Temperature",
    shape = "Clone",
    title = "Clone reaction norms in timing-growth space"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.background = element_rect(fill = "white"),
    legend.position = "right"
  )
# --------------------------------------------------
# 5. bottom panel
#    - clone-level delta plot
#    - color by leaf
# --------------------------------------------------
p2 <- ggplot(delta_df, aes(x = delta_t, y = delta_mu, color = leaf_id)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4) +
  geom_point(size = 2.8, alpha = 0.95) +
  geom_smooth(
    aes(group = 1),
    method = "lm",
    se = FALSE,
    linewidth = 0.7,
    color = "black"
  ) +
  labs(
    x = expression(Delta * t[mu[max]] ~ "(30" * degree * "C - 25" * degree * "C)"),
    y = expression(Delta * mu[max] ~ "(30" * degree * "C - 25" * degree * "C)"),
    color = "Leaf",
    title = expression("Clone-level change in timing vs. change in " * mu[max])
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    legend.position = "right"
  )

# --------------------------------------------------
# 6. combine
# --------------------------------------------------
combined_plot <- p1 / p2 +
  plot_annotation(tag_levels = "A")

combined_plot


ggplot(growth_summary, aes(x = t_mumax, y = mu_max, color = Treatment)) +
  geom_point(alpha = 0.6) +
  
  geom_smooth(
    method = "lm",
    se = FALSE,
    linewidth = 1
  ) +
  
  labs(
    x = expression(t[mu[max]]),
    y = expression(mu[max]),
    color = "Temperature",
    title = "Relationship between timing and growth rate across temperatures"
  ) +
  theme_bw() +
  theme(panel.grid = element_blank())

ggplot(growth_summary, aes(x = t_mumax, y = mu_max)) +
  geom_point(aes(color = Treatment), alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_bw()


ggplot(clone_means, aes(x = t_mumax, y = mu_max, color = Treatment)) +
  
  # clone reaction norms (thin lines)
  geom_line(aes(group = clone_id), alpha = 0.4) +
  
  # clone mean points
  geom_point(size = 2.5) +
  
  # regression lines with confidence bands
  geom_smooth(
    method = "lm",
    se = TRUE,
    linewidth = 1,
    alpha = 0.2
  ) +
  
  labs(
    x = expression(t[mu[max]]),
    y = expression(mu[max]),
    color = "Temperature"
  ) +
  theme_bw() +
  theme(panel.grid = element_blank())
```
